---
title: An R Markdown document converted from "Section 2_2.ipynb"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# 2.2 Detrending, Differencing and Variance Stabilization

-   As we mentioned before, the sample mean or sample
    autocovariance/correlation is a meaningful estimate of a theoretical
    counterpart, only if the time series is weakly stationary.

-   Often, real-life time series data exhibits non-stationarity.

-   We will mention some methods playing down the effects of
    nonstationarity so the stationary properties of the series may be
    studied.

**Note:**

-   The idea of transforming "non-stationarity" to "stationarity" is NOT
    the only way to start a time series analysis. Some methologies may
    directly tackle "non-stationarity" (see, e.g., the project topics
    Unit Root, State Space Models).

-   Yet the strategy underlies a classical approach towards time series
    analysis known as the [Box–Jenkins
    Approach](https://en.wikipedia.org/wiki/Box%E2%80%93Jenkins_method),
    and can also be viewed as a useful exploratory step in time series
    analysis.

```{r}
#install.packages('astsa'); 
library('astsa')
```

```{r}
# Examples of non-stationary time series data
plot(jj, type="o", ylab="Quarterly Earnings per Share")
plot(gtemp_land, type="o", ylab="Global Land Temperature Deviations")

```

## 2.2.1 Trend Stationarity and Detrending

The arguably easiest form of non-stationarity to work with is the
**trend stationary**.

**Definition** A time series model of $\{x_t\}$ is said to be **trend
stationary**, if it takes the form $$
x_t=\mu_t+ y_t
$$ where $x_t$ are the observations, $\mu_t$ is the (random or
nonrandom) trend, $y_t$ is a weakly stationary process.

If the trend $\mu_t$ is regular (e.g., smooth, even following an
algebraic form), a natural approach to proceed (**detrend**):

1.  Construct a reasonable estimate of the trend component $\mu_t$, say
    $\hat{\mu}_t$.

2.  Work with the residuals $\hat{y}_t = x_t -\hat{\mu}_t$ as an
    approximation of $y_t$.

**Example**. Detrending Chicken Prices

Recall in the previous section, we essentially worked with
$$\mu_t=\beta_0+\beta_1 t $$ (here $t$ is in the unit of year, which is
identical to $z_t$ in the previous section).

```{r}
plot(chicken)
summary(fit <- lm(chicken~time(chicken), na.action=NULL)) # regress chicken on time
plot(resid(fit), type="o", main="detrended")
acf(resid(fit), 48, main="detrended")
```

## 2.2.2 Differencing and Stationarity

**Observation**: chicken price data seems to exhibit random trends (also
present in the residuals).

An alternative model: in the trend stationary model $$
x_t=\mu_t+ y_t, \tag{1}
$$ we model $\mu_t$ as random walk with drift: (below let $t$ be
integers) $$
\mu_t= \delta+ \mu_{t-1} +w_t \tag{2},
$$ where $\{w_t\}$ white noise. Assume $\{w_t\}$ and $\mu_0$ (and hence
$\{\mu_t\}$) are independent of the weakly stationary $\{y_t\}$.

If we follow the model assumptions (1) and (2), then the differenced
series is $$
x_{t}-x_{t-1}=(\mu_t+ y_t) -(\mu_{t-1}+ y_{t-1})= (\mu_t-\mu_{t-1})+(y_t-y_{t-1})=\delta+w_t+ y_t-y_{t-1}.
$$

Exercise 25. Show that the differenced series above is weakly
stationary.

**Example.** Differencing Chicken Price.

```{r}
plot(diff(chicken), type="o", main="differenced")
acf(diff(chicken), 48, main="differenced")
```

If the trend stationary model with the random walk with drift $\delta$
is assumed (namely (1) and (2) above), then one can use the sample mean
of the differenced time series $\{x_{t}-x_{t-1}\}$ to estimate the drift
$\delta$.

```{r}
#par(mfrow=c(2,1))
plot(diff(chicken), type="o")
(delta=mean(diff(chicken))) # drift estimate
acf(diff(chicken), 48)
```

If in the trend stationary model $x_t=\mu_t+ y_t$, we replace the random
walk with drift by the linear trend: $\mu_t=\beta_0+\beta_1t$, then $$
x_t−x_{t−1}= (\mu_t-\mu_{t-1})+(y_t-y_{t-1})= \beta_1+ (y_t-y_{t-1}).
$$ One can show similarly as above that the differenced series is weakly
stationary with mean $\beta_1$.

The so under this linear trend model, the sample mean of
$\{x_t-x_{t-1}\}$ can be viewed as an estimate of $\beta_1$.

Exercise 24.

```{r}
fit$coefficient[2]/12 # beta_1 estimate (rescaled for monthly time unit)
delta
```

Exercise 26\*(Optional). Which is a better estimate?

**Observations.**

-   In the trend stationary model, when the trend is linear or (drifted)
    random walk, differencing produces a weakly stationary series.
-   However, the resulting weakly stationary series is not the original
    stationary series $\{y_t\}$ assumed in the model, but
    $\{y_t-y_{t-1}\}$ instead.
-   Hence if the interest is in statistical analysis of $\{y_t\}$ (e.g.,
    estimation of ACF), differencing will obscure the analysis.

```{r}
w=rnorm(1000)
plot(y<-as.ts(w+0.01*(1:1000))) # sample ACF of linear trend plus white noise
acf(w)
acf(diff(y)) # sample ACF of differenced white noise
```

Comparison: **Detrending** vs **Differencing**:

-   They both can be used to remove trend.
-   Differencing requires no statistical estimation, whereas detrending
    typically needs one.
-   If *trend stationary model* is assumed and analysis of $\{y_t\}$ is
    essential, then detrending may be more appropriate.
-   If the goal is to coerce the data to stationarity, then differencing
    may be more appropriate.

## 2.2.3 Backshift and Differencing Operators

One can difference a differenced series again.

```{r}
#install.packages('astsa'); 
library('astsa')
y <- diff(chicken)
plot(y, main="first difference")
acf(y,48)
z<- diff(y)
plot(z, main="second difference")
acf(z,48)
```

Let's check how the algebra of repeated differencing works: suppose
$\{x_t\}$ is the observed time series. After a differencing, we get the
differenced series $\{y_t=x_{t}-x_{t-1}\}$. Now apply another round of
differencing:

$$
z_t=y_t-y_{t-1}=(x_{t}-x_{t-1}) -(x_{t-1}-x_{t-2})= x_t-2x_{t-1}+x_{t-2}.
$$

Even another: $$
u_t=z_{t}-z_{t-1}=(x_t-2x_{t-1}+x_{t-2}) -(x_{t-1}-2x_{t-2}+x_{t-3})=x_t -3 x_{t-1} +3x_{t-2} -x_{t-3}.
$$

Question: What do these coefficients $(1,-2,1)$ and $(1,-3,3,-1)$ look
like?

To facilitate the algebra of differencing, and in general linear
filtering of time series, we introduce the followong concept.

**Definition.** Given a time series $\{x_t\}$, we define the **backshift
operator** $B$ as the operation $$B x_{t} =x_{t-1},$$\
and the identity operator $I$ as the operation $$I x_t= x_t.$$

Remark: like for linear filters, mathematically one can view $B$ and $I$
and operators which takes a series as an input and yields another series
as an output. Indeed, $B$ is like the *lag* function in R:
*lag(x,k=-1)*, which shifts the time index by -1.

**Definition.** Given a time series ${x_t}$, we define the forward-shift
operator $B^{-1}$ as the operation $$
B^{-1} x_{t} =x_{t+1},
$$\
and $k$-times iterated backshift and forward-shift operators as $$
B^k x_t = B(\dots   B(B x_t))=x_{t-k},   \quad B^{-k} x_t = B^{-1}(\dots   B^{-1}(B^{-1} x_t))=x_{t+k},\quad k=1,2,3,\ldots,
$$ and set $$
B^0=I \quad(\text{sometimes as }1).
$$

**Note.**\
$BB^{-1}=B^{-1}B=I$. $B^m B^n=B^{m+n}$ for any integers (positive,
negative, or zero) $m$ and $n$.

Any linear filter
$\{x_t\}\rightarrow \{\sum_{k}a_{t-k} x_k=\sum_{k} a_{k} x_{t-k}\}$ can
be expressed using the shift operators as $$
\left(\sum_{k} a_k B^k\right) x_t =\sum_{k} a_{k} B^kx_{t}=\sum_{k} a_{k} x_{t-k}.
$$

For example, the difference operation (a special case of linear filter)
can be expressed as $(I-B)$: $$
(I-B) x_t= x_t-x_{t-1}.
$$

**Definition.** Denote the difference operator as $\nabla=I-B$, we
define the $d$-times differencing operator as $$
\nabla^d x_t=  \nabla (\ldots  \nabla(\nabla x_t)).
$$

For example,
$\nabla^2 x_t= \nabla (x_{t}-x_{t-1}) = x_{t}-2x_{t-1}+ x_{t-2}$.

**Fact.** When compositing multiple layers of linear filters, one can
apply the same algebra rule as the multiplication of polynomials by
treating $B$ as a variable and treating $I$ as "one".

For example, with coefficients $a,b,c,d$, we have $$
(aI+b B)(cI+d B)=  a I \cdot cI + a I \cdot d B + b B \cdot c I+ bB \cdot d B= acI + (ad+bc) B+ bd B^2.
$$

As a special case of applying the general rule above, one can expand
$\nabla^d = (I-B)^d$ with [Binomial
Theorem](https://en.wikipedia.org/wiki/Binomial_theorem) as $$
\nabla^d = \sum_{k=0}^d {d\choose k} (-1)^{d-k} B^{d-k}.
$$

**Note.** The language and algebra of shift operators will be used
extensively in the next Chapter on ARIMA models.

```         
diff(x, lag = 1, differences = 1, …)

differences
an integer indicating the order of the difference.
```

```{r}
t=1:100
plot(x<-as.ts(t^3-150*t^2+5900*t+45000)) # x stores a polynomial of degree 3
```

```{r}
plot(diff(x, differences = 1), main = 'first difference')
plot(diff(x, differences = 2), main = 'second difference')
plot(diff(x, differences = 3), main = 'third difference')
```

**Fact**: $d$-times differencing reduces a polynomial trend of degree
$d$ to a constant trend. (Further differencing will yield zeros.)

Exercise 27.

**Remark.**\* An alternative to differencing is a more flexible
operation called *fractional differencing*. It extends the notion of the
difference operator $(I-B)^d$ to fractional (non-integer) $d$. This is
related to the project topic **long memory** time series.

## 2.2.4 Variance Stabilization

**Example.** Glacial Varve Thicknesses.

Varve: an annual layer of sediment (proxies for climatic parameters such
as temperature, e.g., the warmer the thicker)

![](https://drive.google.com/uc?export=view&id=1uwdbc-cU3mc4Ld20ZGKj4sfHpWxnm3ft)

```{r}
#options(repr.plot.width=20, repr.plot.height=5)
plot(varve, main="varve", ylab="")
plot(log(varve), main="log(varve)", ylab="" )
```

The variability has been stabilized and more "normal".

Exercise 28. Explore more.

A more general family of transforms: Box–Cox. See equation (2.34) in the
textbook.
