---
title: An R Markdown document converted from "Section 2_3.ipynb"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# 2.3 Smoothing in the Time Series Context

Often at the stage of exploratory analysis of time series, one does not
wish to assume a specific parametric form for the mean function of a
time series. Instead, one may want to freely explore traits in a time
series such as long-term trend and seasonal components. For such a
purpose, time series smoothing techniques play important roles.

Broadly speaking, the techniques to be discussed in this section belong
to the topic [nonparametric
statistics](https://en.wikipedia.org/wiki/Nonparametric_statistics).

## 2.3.1 Moving Average Smoother and Seasonality Filtering

Previously, we discussed linear filter

$$
\{x_t\} \longrightarrow  \left\{m_t=\sum_{j} a_{j}x_{t-j}=\left(\sum_{j}a_jB^j\right) x_t \right\} \tag{1}.
$$

We looked at the 3-term moving average $$
\frac{1}{3}(x_{t-1}+x_{t}+x_{t+1}), \tag{2}
$$ which is a special case of linear filter by setting
$a_{-1}=a_0=a_{+1}=\frac{1}{3}$ and $a_j=0$ for $|j|>1$. We discovered
that this moving average has an effect of smoothing out the time series.

We can generalize the moving average in (2) by allowing unequal weights
and more terms, namely, we consider the linear filter (1) with
coefficients $\{a_j\}$ satisfying $$
a_j\ge 0,\quad \sum_{j} a_j =1.
$$

**Remark.**

-   Often the coefficients are chosen such that $a_{-j}=a_j$ (symmetric
    weighting).

-   In a theoretical consideration, one may allow infinitely many
    nonzero $a_j$'s, although in a practical computation, the sum in (1)
    is always a truncated finite sum.

Consider the SOI time series.

```{r}
#install.packages('astsa');
library('astsa')
```

```{r}
#options(repr.plot.width=15, repr.plot.height=6)
acf(soi,lag.max = 48)
```

THe ACF plot suggests possible yearly cycle pattern in the time series.

Let's consider a symmetric moving average ($t$ takes integer values and
is in the unit of month) which corresponds to a yearly cycle: $$
m_t= \frac{1}{24}x_{t-6}+\frac{1}{12}x_{t-5}+\frac{1}{12}x_{t-4}+\ldots+ \frac{1}{12}x_{t+4}+\frac{1}{12}x_{t+5}+ \frac{1}{24}x_{t+6} =\sum_{j} a_j x_{t-j}
$$ with $a_j=\frac{1}{12}$ if $|j|\le 5$, $a_6=a_{-6}=\frac{1}{24}$ and
$a_j=0$ for $|j|>6$.

**Note.** Since the period 12 is an even number, whereas a symmetric
moving average always has an odd number of terms, we have thus placed
half weights to the two boundary points as a compromise.

```{r}
wgts = c(.5, rep(1,11), .5)/12 # weight coefficient vector

nwgts = as.ts(c(rep(0,10), wgts, rep(0,10))) # make a copy of wgts as a ts object allowing time shifts (add some zeros on both sides).
plot(nwgts, type="l", main = 'filter shape', xlab = 'index', ylab = 'coefficient value') # display the shape of the filter coefficients
```

Mathematical thinking behind: If a series $s_t$ has periodicity $12$,
then what would happen for the filtered series: $$
\sum_{j} a_{j} s_{t-j}
$$ with $\{a_j\}$ as above?

```{r}
t=1:200

s=as.ts(sin(2*pi/12*t+1)) # a signal of periodicity 12

plot(s)
lines(filter(s, sides=2, filter=wgts), lwd=3, col=4)
```

Consider the following decomposition $x_t = s_t + y_t$ where $y_t$
captures the fluctuation beyond the periodic component (e.g., a linear
trend and stationary noise) $$
\sum_{j} a_{j} x_{t-j}= \sum_{j} a_{j} s_{t-j} + \sum_{j} a_{j} y_{t-j} = constant  + \sum_{j} a_{j} y_{t-j}.
$$

**Observation.** $\sum_{j} a_{j} x_{t-j}$ will not show periodic
fluctuation (say periodic/seasonal fluctuaion has been filtered out).

**Question**: if $z_t=\beta_0 +\beta_1 t$ is a straight line, what
happens if we apply the same filter $\sum_{j} a_{j} z_{t-j}$ as above?

```{r}
z=as.ts(0.1*t+2) # a linear trend

plot(z)
lines(filter(z, sides=2, filter=wgts), lty=2 ,lwd=3, col=4)
```

Moving average smoother applied to trend+seasonal+noise data.

```{r}
n=length(s)
x=s+0.01*(1:n)+0.5*rnorm(n)

plot(x)
lines(filter(x, sides=2, filter=wgts), lwd=3, col=4)
```

**Summary**: when a (linear trend + seasonal component + noise) model
passes through the moving average smoother, the linear trend is
invariant, seasonal component is filtered out, and the noise is
smoothed.

**Example.** Apply moving average smoother to SOI time series.

```{r}
#plot time series and the moving average

soif = filter(soi, sides=2, filter=wgts)

plot(soi)
lines(soif, lwd=3, col=4)

```

The fluctuation left highlights El Nino effect.

## 2.3.2 Kernel Smoothing

Recall moving average smoother can be represented by a
rectangle/trapezoid "bump". One can generalize such a "bump by
introducing a **kernel** (function): $K(x)$ (controls shape of bump) and
a **bandwidth** $b$ (controls width of bump).

Define the smoother $$
m_t=\sum_{i=1}^n w_i(t) x_i.
$$ where $$
w_i(t)=\frac{K(\frac{t-i}{b})}{\sum_{j=1}^n K(\frac{t-i}{b})}.
$$

Note that $\sum_{i=1}^n w_i(t)=1$ for any $t$, and $m_t$ is defined even
for $t$ outside the sampling time range (e.g., non-integers).

It is common to use a symmetric probability density function for the
kernel $K$.

A common choice of $K$ is the Gaussian kernel (probability density of
normal distribution).

The moving average smoother approximately corresponds to the "box"
kernel $K(x)=1$ for $x\in [-1/2,1/2]$, and $K(x)=0$ otherwise.

**Remark.**

-   In R kernel smoothing is implemnted by the function *ksmooth*:

```         
ksmooth(x, y, kernel = c("box", "normal"), bandwidth = 0.5,
        range.x = range(x),
        n.points = max(100L, length(x)), x.points)

x
input x values. Long vectors are supported.

y
input y values. Long vectors are supported.

kernel
the kernel to be used. Can be abbreviated.
All kernels are scaled so the upper and lower quartiles of the kernel (viewed as a probability density) are +/- 0.25.

bandwidth
the bandwidth.
```

-   In nonparametric statistics, kernel smoother is also called a
    Nadaraya-Watson estimator, used to estimate the smoothly varying
    mean.

-   Often, the choice of bandwidth $b$ is much more important than the
    choice of the kernel $K$.

-   As described above, the symmetric kernel $K$ used in R is scaled so
    that $\int_{-0.25}^{0.25} K(x)dx=0.5$. (For Gaussian kernel this is
    the same as the density of $N(0,0.37^2)$.)

-   We will not elaborate on how $b$ should be chosen. It can be chosen
    in a data-driven way (e.g., cross-validation), which may be
    discussed in a nonparametric statistics course. In practice it is
    useful to try different $b$.

-   Check
    [KernSmooth](https://cran.r-project.org/web/packages/KernSmooth/KernSmooth.pdf)
    package for modern implementation with choice of $b$ included.

How does kernel smoothing affect periodic and linear trend?

```{r}
t=1:200
s=as.ts(sin(2*pi/12*t+1)+1) # a signal of periodicity 12

plot(s)
lines(ksmooth(x = t, y = s, kernel = "normal", bandwidth = 5), lwd=3, col=4)
```

```{r}
z=as.ts(0.1*t+2) # a linear trend

plot(z)
lines(ksmooth(x = time(z), y = z, kernel = "normal", bandwidth = 150), lwd=3, col=4)
```

Different bandwidth choices may reveal different features of trends.

```{r}

t=1:200
s=as.ts(sin(2*pi/12*t+1)+1) # a signal of periodicity 12
n=length(s)
x=s+0.01*(1:n)+0.5*rnorm(n)

plot(x) #black: signal + noise
lines(ksmooth(x = time(x), y = x, kernel = "normal", bandwidth = 5), lwd=3, col=4) #blue: smoother
lines(s+0.01*(1:n),lwd=1,col=2)# red: true signal (trend)
```

**Example.** Apply kernel smoothing to SOI.

```{r}
plot(soi, main = 'Gauss Kernel')
lines(ksmooth(x = time(soi), y = soi, kernel = "normal", bandwidth = 4), lwd=2, col=4)
 
 

 

plot(soi, main = 'Box Kernel')
lines(ksmooth(time(soi), soi, "box", bandwidth=2), lwd=2, col=4)

 
 
```

## 2.3.3 LOWESS (locally weighted estimated scatterplot smoothing) / Local regression

By Taylor expansion, we know that a smooth function can be *locally*
approximated by a linear function, or more generally, a polynomial.

This idea motivates us to consider *locally* fit a linear line /
polynomial to time series.

LOWESS / Local regression is an approach which implements this idea.

-   A precise description of
    [LOWESS](https://en.wikipedia.org/wiki/Local_regression) will be
    discussed in an advanced nonparametric statistics course. Roughly
    speaking, LOWESS performs a polynomial regression by downweighting
    the points that are far from the time point of interest (similar
    idea to kernel smoothing).

-   LOWESS also involves a choice of bandwidth, reformulated as the
    *smoother span* in R, which controls strength of smoothing.

-   The R default choice of degree of polynomial = 1. In fact, kernel
    smoothing is approximately equivalent to LOWESS with degree = 0.

-   Check
    [loess](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/loess),
    a newer formula based version of *lowess*.

```         
lowess(x, y=NULL, f=2/3, iter=3L, delta=0.01 *
       diff(range(x)), ...)
x, y
vectors giving the coordinates of the points in the scatter plot. Alternatively a single plotting structure can be specified.

f
the smoother span. This gives the proportion of points in the plot which influence the smooth at each value. Larger values give more smoothness.
```

```{r}
t=1:200
z=as.ts(0.1*t+2) # a linear trend
plot(z)
lines(lowess(z,f=0.1), lty=2, lwd=2, col=2)
```

```{r}
u=as.ts(t^3-250*t^2+5900*t+45000) # a cubic trend
plot(u)
lines(lowess(u,f=0.1), lty=2, lwd=2, col=2)

plot(u)
lines(ksmooth(x = t, y = u, kernel = "normal", bandwidth = 1000), lwd=3, col=4)

# big bandwidth comparison of kernel smoothing vs LOWESS
```

```{r}
x=u+rnorm(200,sd=80000) # a cubic trend plus noise
plot(x)
lines(u,col = 3)
lines(lowess(x,f=0.1), lty=2, lwd=2, col=2)
```

**Observation**:

-   When smoothing span (bandwidth) is properly chosen, LOWESS can be
    used to uncover a smooth trend.

-   The trend fluctuation may also be dampened by LOWESS if smoothing
    span is too large.

-   When using LOWESS to uncover noisy trend, there is a tradeoff in
    choosing smoothing span: too large $\rightarrow$ trend dampened; too
    small $\rightarrow$ too much noise retained

```{r}
s=as.ts(sin(2*pi/100*t+1)+1) # a fluctuating signal
x=s+rnorm(200,sd=0.2)
```

```{r}
plot(x)

lines(lowess(x,f=0.1), lty=2, lwd=2, col=2) #LOWESS smoother
```

**Example.** Apply LOWESS to SOI.

```{r}
plot(soi)
lines(lowess(soi, f=.05), lwd=2, col=4) # attempt to capture El Nino cycle
lines(lowess(soi), lty=2, lwd=2, col=2)
```

## 2.3.4 Smoothing Splines

There is another popular nonparametric smoothing technique called
[smoothing splines](https://en.wikipedia.org/wiki/Smoothing_spline). The
mathematical motivation of the technique is somewhat different from the
local averaging/fitting techniques we have seen above.

Suppose the observed time series is $\{x_1,x_2,\ldots,x_n\}$ and we want
to fit a smoother $m_t$. For mathematical convenience, let's treat $t$
as a real number which can take non-integer values.

The least squares criterion for finding best fit is to minimize

$$
SSE=\sum_{t=1}^n(x_t-m_t)^2.
$$

When $m_t$ is in a parametric form, e.g., staight line
$m_t=\beta_0+\beta_1 t$, we find parameters, e.g., $\beta_0$ and
$\beta_1$ which minimize SSE.

Now suppose we do not wish to impose a parametric form for $m_t$,
namely, we want to find a general curve which minimizes SSE above.

However, if we do not have any additional constraint about the general
curve, the problem is ill-posed.

Question: why?

To make the question well-posed, introduce an additional penalty term
into the criterion:

$$
Criterion=SSE+Penalty= \sum_{t=1}^n(x_t-m_t)^2 + \lambda \int (m_t'')^2 dt,
$$ where $\lambda>0$ is a parameter chosen by user.

-   Mathematically it can be shown that the curve $m_t$ which minimizes
    the criterion above is a cubic
    [spline](https://en.wikipedia.org/wiki/Spline_(mathematics)),
    piecewise polynomials with degrees $\le 3$, which can be efficiently
    numerically evaluated.

-   The penalty integral $\lambda \int (m_t'')^2 dt$ measures the "total
    curvature" of the curve ($m_t''=0$ corresponds to straight line).
    The larger the penalty parameter $\lambda$ is, the harder it makes
    for the fitted line to "turn", and hence $\lambda$ controls the
    smoothness (Note that $\lambda=\infty$ corresponds to fitting
    straight line with OLS.)

(For an alternative physical interpretation, see textbook.)

-   As $\lambda\downarrow 0$, the solution curve approaches the
    [interpolating
    spline](https://en.wikipedia.org/wiki/Spline_interpolation), a curve
    that goes through every $(t,x_t)$ point.

In R the smoothing spline is implemented by the function
[smooth.spline](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/smooth.spline).

```         
smooth.spline(x, y = NULL, w = NULL, df, spar = NULL, lambda = NULL, cv = FALSE,
              all.knots = FALSE, nknots = .nknots.smspl,
              keep.data = TRUE, df.offset = 0, penalty = 1,
              control.spar = list(), tol = 1e-6 * IQR(x), keep.stuff = FALSE)

x
a vector giving the values of the predictor variable, or a list or a two-column matrix specifying x and y.

y
responses. If y is missing or NULL, the responses are assumed to be specified by x, with x the index vector.

spar
smoothing parameter, typically (but not necessarily) in \((0,1]\). Alternatively lambda may be specified instead of the scale free spar=\(s\).

if smoothing parameter is not specified, cross validation is used to choose smoothing parameter
```

```{r}
t=1:200
u=as.ts(t^3-250*t^2+5900*t+45000) # a cubic trend
plot(u)
lines(smooth.spline(time(u), u, spar=0.999), lwd=2, lty=2, col=4)
```

```{r}
s=as.ts(sin(2*pi/100*t+1)+1) # a fluctuating trend
plot(s)
lines(smooth.spline(time(s), s, spar=0.95), lwd=2, lty=2, col=4)
```

```{r}
x=s+rnorm(200,sd=0.2) # signal plus noise
plot(x)
lines(smooth.spline(time(x), x, spar=0.8), lwd=2, lty=2, col=4)
```

**Example.** Apply Smoothing Spline to SOI.

```{r}
plot(soi)
lines(smooth.spline(time(soi), soi, spar=.5), lwd=2, col=4) # spar is a log rescaled version of lambda
lines(smooth.spline(time(soi), soi, spar= 1), lty=2, lwd=2, col=2)
```

```{r}
plot(time(soi),soi)
lines(smooth.spline(time(soi), soi), lty=2, lwd=2, col=2) # if smoothing parameter is not specified, cross validation is used to choose smoothing parameter
```

## 2.3.5 Smoothing Scatterplot

The smoothing techniques discussed so far are essentially applied to
data pairs $\{(1,x_1),\ldots, (n,x_n)\}$. If one has another observed
time series $\{y_1,\ldots,y_n\}$, one can also apply the smoothers to
the pairs $\{(x_1,y_1),\ldots, (x_n,y_n)\}$ to explore the marginal
dependence structure.

**Example.**

```{r}
options(repr.plot.width=10, repr.plot.height=10)
plot(tempr, cmort, xlab="Temperature", ylab="Mortality")
lines(lowess(tempr, cmort, f=0.2))
```

One can also apply the smoothing technique to explore lagged-value
scatterplot.

*astsa* includes functions such as
[lag1.plot](https://www.rdocumentation.org/packages/astsa/versions/1.14/topics/lag1.plot)
(single time series) and
[lag2.plot](https://www.rdocumentation.org/packages/astsa/versions/1.14/topics/lag2.plot)
(cross time series) for lagged-value scatterplot with LOWESS smoother.

```{r}
options(repr.plot.width=15, repr.plot.height=15)
lag1.plot(soi, 3)
```

```{r}
lag2.plot(soi, rec, 3)
```

**Summary.**

-   Local average/regression based methods: Moving Average, Kernel
    Smoother, LOWESS.

-   Smoothing Spline.

-   Typically, the performance of these smoothing techniques are
    similar. The more important choice is the strength of smoothing.

-   With a proper choice of strength of smoothing, the smoothing curve
    can be viewed as a nonparameteric estimate of the underlying smooth
    trend.

Exercises 29, 30

Certain smoothing techniques can also be used to produce forecast.

For example, [exponential
smoothing](https://en.wikipedia.org/wiki/Exponential_smoothing).

Project topic: Forecasting with ETS.

```{r}
fit <- HoltWinters(chicken,alpha=0.5, beta=FALSE, gamma= FALSE) #exponential smoothing
pred<-predict(fit, n.ahead =10, prediction.interval = TRUE)
plot(fit,pred)
```
