---
title: An R Markdown document converted from "Section 1_4.ipynb"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# 1.4 Stationary Time Series

## 1.4.1 Univariate Case

In statistical modeling, we look for an *invariant* mathematical rule
which captures the entire data set.

For example, in many contexts, we assume the data is generated
(independently) from the same distribution.

In the context of time series, an important notion of invariance is
*stationarity*.

Roughly speaking, the notion stationarity says the time series is
generated by the same (random) mechanism in a time-invariant manner.

**Definition**\* A time series (model) $\{x_t\}$ is said to be
**strictly stationary**, if the joint distribution of
$(x_{t_1}, x_{t_2},\ldots, x_{t_k})$ is identical to that of
$(x_{t_1+\delta}, x_{t_2+\delta},\ldots, x_{t_k+\delta})$, for any
$k\in \mathbb{Z}_+$, any time indices $t_1,\ldots, t_k$ and any shift
$\delta$. In other words, all the finite-dimensional joint distributions
are invariant under time shift.

**Remark**:

-   In words, stationarity means all the statistical characteristics of
    the time series is time-invariant. (For example, the sample obtained
    from time period 10am to noon is the equivalent to that obtained
    from 2pm to 4pm)

-   Assumption "i.i.d." $\Rightarrow$ Strict stationarity.

Strict stationarity relaxes the first "i", i.e., the independence
assumption, and generalizes the second "i", i.e., identical distribution
assumption.

```{r}
 # Some examples of strictly stationary time series
 #par(mfrow=c(2,1))
 plot(w<-as.ts(rnorm(1000)), main="Gaussian white noise")
 plot((filter(w, sides=2, filter=rep(1/3,3))), main = "Moving Average")

```

```{r}
#Some examples of non-stationary time series
#par(mfrow=c(3,1))


cs = 2*cos(2*pi*1:500/50 + .6*pi); w = rnorm(500,0,1)
plot.ts(cs+w, main="signal plus noise")

plot.ts(w*seq(1,5,length.out=500),main='noise of increasing variance')

plot.ts(filter(rnorm(500) , filter=1 , method="recursive") , main="random walk")
```

Visual features of realized stationary time series:

-   Trendless
-   Constant variability
-   Constant smoothness/roughness

**Important consequences of strict stationarity assumption**:

Suppose $\{x_t\}$ is strictly stationary and its mean and variance
exist. Then the following holds.

1.  The mean function $\mu(t)=\mu$ for some constant $\mu$.
2.  The autocovariance function $\gamma(s,t)=\gamma(s+\delta,t+\delta)$
    for any possible $s,t,\delta$.

Explanation for 1: all $x_t$, as $t$ varies, have the same marginal
distribution.

Explanation for 2: $(x_s,x_t)$ has the same bivariate distribution as
the time shifted $(x_{s+\delta},x_{t+\delta})$.

**Note.**

-   Property 2 says $\gamma(s,t)=\gamma(t,s)$ depends only on $|t-s|$,
    not on the positions of $s$ and $t$.

-   Property 2 also implies that the variance $Var(x_t)=\gamma(t,t)$ is
    a constant.

**Definition.** A time series $\{x_t\}$ is said to be **weakly
stationary**, if Properties 1 and 2 above hold.

**Note.** Obviously, strictly stationarity (with existence of mean and
variance) $\Rightarrow$ weak stationarity.

Weak stationarity is about time-invariance in mean and covariance
structures, while strict stationarity about time-invariance in all
possible statistical aspects.

**Definition.** Under weak stationarity, we also (and more commonly)
call the univariate function
$$\gamma(h):=\gamma(t+h,t) \text{ (the same value for any $t$)} $$ the
**autocovariance function**, and call the univariate function
$$\rho(h):=\gamma(h)/\gamma(0)$$ (if $\gamma(0)\neq 0$) the
**autocorrelation function (ACF)**.

A useful relation between bivariate and univariate autocovariance
functions of weakly stationary time series:

$$
\gamma(s,t)=\gamma(s-t)=\gamma(t,s)=\gamma(t-s).
$$ So $$
\gamma(h)=\gamma(-h).
$$

Often the concept *weak stationarity* is easier to handle compared with
*strict stationarity*, although the latter is important for certain
theoretical considerations.

\
**(The textbook eventually uses "stationary" to mean "weakly
stationary". For homework problems in the textbook involving justifying
that a time series is "stationary", it typically means "weakly
stationary" unless specified otherwise.)**

**Definiton.**\* A time series is said to be a Gaussian process, if all
finite-dimensional distributions are multivariate normal.

**Fact**\*: For Gaussian processes, mean and covariance structures
completely characterize its properties. So weak stationarity and strict
stationarity are equivalent for Gaussian processes.

**Example.** A white noise series $\{w_t\}$ is weakly stationary. The
autocovariance and autocorrelation function is given by $$
\gamma(h)=
\begin{cases}
\sigma_w^2 & h=0,\\
0   &h\neq 0,
\end{cases},
\qquad
\rho(h)=
\begin{cases}
1 & h=0 , \\
0   &h\neq 0.
\end{cases}
$$

(On the other hand, a white independent noise series is strictly
stationary.)

**Example.** Let $\{w_t\}$ be white noise and consider
$x_t=\frac{1}{3}(w_{t-1}+w_t+w_{t+1})$. In the previous section, we have
derived its mean function $$
\mu(t)=0.
$$ We have also effectively derived the autocovariance function $$
\gamma(h)=  
\begin{cases}
\frac{3}{9}\sigma_w^2  & h=0,\\
\frac{2}{9}\sigma_w^2  & h= \pm 1,\\
\frac{1}{9}\sigma_w^2  & h= \pm 2,\\
0 & |h|>2,
\end{cases}
$$ which yields the ACF $$
\rho(h)=  
\begin{cases}
1  & h=0,\\
\frac{2}{3}   & h= \pm 1,\\
\frac{1}{3}   & h= \pm 2,\\
0 & |h|>2.
\end{cases}
$$

```{r}
plot(-4:4, c(0,0,1/3,2/3,1,2/3,1/3,0,0), type = "h", xlab = "h", ylab = "ACF")
axis(at = -4:4, side=1)
abline(h=0)
```

The autocovariance/correlation functions of weakly stationary time
series has the following properties:

(1) $\gamma(h)=\gamma(-h)$, $\rho(h)=\rho(-h)$ (enough to describe
    $h\ge 0$).

(2) $|\gamma(h)|\le \gamma(0)$, $|\rho(h)|\le \rho(0)=1$ (maximized at
    $h=0$).

(3)\* $\gamma$ or $\rho$ is non-negative definite (see Problem 1.25, NOT
included in homework)).

In fact, one can generalize the example above

**Definition.** Suppose $\{w_t\}$ is white noise with marginal variance
$\sigma_w^2\in (0,\infty)$. Then the time series $\{x_t\}$ defined by
the linear filter and a shift as follows $$
x_t=\mu+\sum_j\psi_jw_{t-j}     \tag{1}
$$ is called a **linear process**, where $\mu$ is a non-random constant,
and $\psi_j$'s are non-random coefficients.

\*Some mathematical care is needed when infinitely many $\psi_j$'s are
nonzero. E.g, people often impose some assumption such as
$\sum_j\psi_j^2<\infty$ or $\sum_j|\psi_j|<\infty$.

One can verify that a linear process as above is weakly stationary with
mean function $$
\mu(t)=E(x_t)=\mu+ \sum_j\psi_j E(w_{t-j})=\mu
$$ and autocovariance function\
$$
\gamma(t+h,t)=\sigma_w^2 \sum_j \psi_{j+h}\psi_j=\gamma(h).
$$

Exercise 13. Derive the autocovariance function above.

**Definition.** If the linear process in (1) satisfies $\psi_{j}=0$ for
all $j<0$, then the linear process is said to be *causal*.

**Wold Decomposition**\*: under cetain mild condition, a weakly
stationary time series can always be expressed as a causal linear
process. (Theorem B.5 in the textbook.)

Next, we turn to an example which violates stationarity.

**Example.** A random walk $x_t=w_1+\ldots + w_t$, is NOT weakly
stationary.

*Argument 1*: its (bivariate) autocovariance function $$
\gamma(s,t)=\sigma_w^2\min(s,t)
$$ depends the specific location of $s$ and $t$, not only on $|s-t|$;

*Argument 2* (simpler): the variance $Var(x_t)=\gamma(t,t)=\sigma_w^2 t$
is NOT a constant with respect to time index $t$.

(However, the mean function $\mu(t)=0$, from which we cannot find a
violation of stationarity.)

```{r}
s <- t <- seq(1, 20)
gamma=outer(s,t,pmin) # compute minimum(s,t) over a mesh of (s,t)
persp(s, t, gamma,   theta = -20, phi = 20) # plot the bivariate autocovariance function (with white noise variance 1), theta and phi are spatial angles
```

Exercise 12. Explore the relation between stationarity and trend.

## 1.4.2 Multivariate Case

The discussions so far extend naturally to multivariate time series as
well.

**Definition** A multivariate time series
$\{\mathbf{x}_t\}=\{(x_{t1},\ldots, x_{tr})^T\}$ is said to be strictly
stationary, if all the finite-dimensional distributions are invariant
under time shift.

**Definition** A multivariate time series
$\{\mathbf{x}_t\}=\{(x_{t1},\ldots, x_{tr})^T\}$ is said to be weakly
stationary, if

1.  The（multivariate) mean function
    $\boldsymbol{\mu}(t)=(E(x_{t1}),\ldots, E(x_{tr}))^T=\boldsymbol{\mu}$
    for some constant vector $\boldsymbol{\mu}=(\mu_1,\ldots, \mu_r)^T$.
2.  The cross covariance functions $\gamma_{jk}(s,t)=Cov(x_{sj},x_{kt})$
    are invariant under time shift, namely,
    $\gamma_{jk}(s,t)=\gamma_{jk}(s+\delta,t+\delta)$ for all possible
    $j,k,s,t,\delta$.

Under the assumptions above, one can introduce the univariate
cross-covariance/correlation functions (depending only on the time
difference $h=s-t$) as before, namely $\gamma_{jk}(h)$ and
$\rho_{jk}(h)$.

**Example.** Recall the bivariate time series $$
(x_t,y_t)= (w_t+w_{t-1}, w_t-w_{t-1}),  \quad \{w_t\} \text{ white noise with }Var(w_t)=\sigma_w^2.
$$ In the previous section, we have derived the mean function $$
(\mu_x(t),\mu_y(t))=(0,0)
$$ and bivariate cross-covariance functions $$
\gamma_{xx}(s,t)=Cov(x_s,x_t)=
\begin{cases}
2\sigma_w^2 &
|s-t|=0\\
\sigma_w^2 & |s-t|=1\\
0 & |s-t|>1
\end{cases},\quad \gamma_{yy}(s,t)=Cov(y_s,y_t)=
\begin{cases}
2\sigma_w^2 &
|s-t|=0\\
-\sigma_w^2 & |s-t|=1\\
0 & |s-t|>1
\end{cases} \quad
\gamma_{xy}(s,t)=Cov(x_s,y_t)=
\begin{cases}
0 &
|s-t|=0\\
\sigma_w^2 & s-t=1\\
 -\sigma_w^2 & s-t=-1   \\
0 & |s-t|>1
\end{cases}
$$

Hence $(x_t,y_t)$ is weakly stationary, and one can derive the
univariate cross-correlation functions as $$
\rho_{xx}(h)=
\begin{cases}
1 &
h=0\\
1/2 & |h|=1\\
0 & |h|>1
\end{cases},\quad
\rho_{yy}(h)=\begin{cases}
1 &
h=0\\
-1/2 & |h|=1\\
0 & |h|>1
\end{cases} \quad
\rho_{xy}(h)=
\begin{cases}
0 &
h=0\\
1/2 & h=1\\
 -1/2 & h=-1   \\
0 & |h|>1
\end{cases}
$$

Exercise 14. Try a similar computation yourself of cross-correlation.

**Note**:\
For weakly stationary multivariate time series,\
cross-covariance (or correlation) satisfies
$$\gamma_{jk}(h)=\gamma_{kj}(-h),$$ whereas $$
\text{typically} \   \gamma_{jk}(h)\neq\gamma_{jk}(-h),\ \text{when } j\neq k.
$$

Exercise 15. Verify the first claim above.

**Example.** Consider a bivariate weakly stationary time series
$(x_t,y_t)$. Suppose we model a time-delayed impact of $\{x_t\}$ on
$\{y_t\}$ by $$
y_t=Ax_{t-\ell}+w_t
$$ where $\ell$ is the time lag, $A$ is a coefficient capturing the
effect, and $\{w_t\}$ is a noise series which is **uncorrelated** with
$\{x_t\}$.

We can compute the cross-covariance as $$
\gamma_{yx}(h)=Cov(y_{t+h},x_t)=Cov(Ax_{t+h-\ell}+w_{t+h},x_t)=Cov(Ax_{t+h-\ell},x_t)=A \gamma_{xx}({h-\ell}).
$$

Recall that $\gamma_{xx}(h-\ell)$ achieves maximum value at $h=\ell$.
Hence one expects to see a peak (may be positive or negative depending
on $A$) of $\gamma_{yx}(h)$ at $h=\ell$ under the model above.

*ts* object

```         
ts(data = NA, start = 1, end = numeric(), frequency = 1,
   deltat = 1, ts.eps = getOption("ts.eps"), class = , names = )
```

Note the *start* and *end* variable. They specify the starting and
ending points of time index, which can make certain time series
operations quite convenient.

```         
lag(x, k = 1, …)

Obtain a lagged version of a time series, shifting the time base back by a given number of observations.
```

```{r}
x = as.ts(c(1,2,3,4,5,4,3,2,1))
y = lag(x,k=-5) # $x lags y by k time units

tsp(x) #returns the tsp attribute (start,end,frequency)
tsp(y)

ts.plot(x,y,lty=1:2) #ts.plot can plot multiple time series by taking into account of start and end
```

```{r}
(z=na.omit(c(NA,1,2,NA))) # a function which removes NA
```

```{r}
x = as.ts(na.omit(filter(rnorm(1002),filter=c(1/3,1/3,1/3)))) #generate x as moving average, drop NA at both ends
y = -2*lag(x, -10) + rnorm(1000) # generate y using the time-lagged impact model above with lag 10
```

```{r}
ts.plot(x, y, col=1:2) # plot x black; y red
```

```{r}
acf(x, ylab= 'ACovF', type = 'covariance') #compute a sample version of  covariance function
ccf(y, x, ylab='CCovF', type= 'covariance') #compute a sample version of cross covariance function (will be elaborated later)
```

```{r}
#install.packages('astsa');
library('astsa')
```

```{r}
acf(soi) # rec: fish population  soi: climate index
ccf(rec,soi)
```
