---
title: An R Markdown document converted from "Section 2_1.ipynb"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

In Chapter 2, we will address multiple linear regression in a time
series context, model selection, exploratory data analysis for
preprocessing nonstationary time series (for example trend removal), the
concept of differencing and the backshift operator, variance
stabilization, and nonparametric smoothing of time series.

# 2.1 Classical Regression in the Time Series Context

We have addressed autoregression models, which is modeling the
relationship between current time series value $x_t$, and the lagged
values $x_{t-1},\ldots, x_{t-q}$. So in a sense, it is modeling an
*internal* evolving mechanism of the time series $\{x_t\}$ itself.

Often, one may also wish to model how a time series $\{x_t\}$ is related
to external factors, say, $(z_{t,1},\ldots, z_{t,q})^T$ (a different
multivariate time series). For example, $\{x_t\}$ is certain pollution
index, $z_{t,1}$ is temperature, $z_{t,2}$ is humidity, etc.

In general, we consider the following multiple linear regression model

$$
x_t=\beta_0+\beta_1 z_{t,1}+\ldots+ \beta_q z_{t,q}+w_t,\tag{1}
$$ where $\beta_0,\ldots,\beta_q$ are unknown coefficients, and
$\{w_t\}$ is random error series with variance $\sigma_w^2$.

**Note.** \* For simplicity and exploratory analysis, we will for now
assume that $\{w_t\}$ is Gaussian white noise independent of
$\{(z_{t,1},\ldots, z_{t,q})^T\}$. However in practice, the assumption
rarely holds, and typically the random error $\{w_t\}$ exhibits temporal
correlation.

-   When the goal is to obtain an accurate estimate of
    $\beta_0,\ldots,\beta_q$, the temporal correlation in $\{w_t\}$ must
    be accounted for, e.g., by using [Generalized Least
    Squares](https://en.wikipedia.org/wiki/Generalized_least_squares).

(\*However, it is known that for some common choice of deterministic
predictor variables such as trigonometric and polynomial functions in
$t$, as sample size tends to infinity, the advantage of GLS over OLS
becomes negligible.)

-   In addition, all the analysis below are conditional on
    $\{(z_{t,1},\ldots, z_{t,q})^T\}$, namely, we are treating these
    variables as if they are nonrandom, although they can be random in
    general.

**Summary: the linear regression analysis to be discussed in this
section which ignores temporal correlation should only be viewed as a
step of exploratory analysis. In particular, do NOT draw conclusions
based on p-values presented in this section. We will revisit this
later.**

## 2.1.1 Least Squares Estimate

It is convenient to write the model (1) with a vector form: introducing
$\mathbf{z}_t=(1,z_{t,1},\ldots, z_{t,q})^T$, and
$\boldsymbol{\beta}=(\beta_0,\beta_1,\ldots,\beta_q)^T$.

$$
x_t= \boldsymbol{\beta}^T \mathbf{z}_t +w_t.
$$

Recall the OLS ([ordinary least
squares](https://en.wikipedia.org/wiki/Ordinary_least_squares))
estimation method finds $\boldsymbol{\beta}$ which minimizes $$
Q(\boldsymbol{\beta})=\sum_{t=1}^n (x_t-\boldsymbol{\beta}^T \mathbf{z}_t)^2.
$$ Denoting the resulting estimate by $\hat{\boldsymbol{\beta}}$, one
can derive that $$
\hat{\boldsymbol{\beta}}=(\hat{\beta}_0,\hat{\beta}_1,\ldots,\hat{\beta}_q)^T= \left(\sum_{t=1}^n \mathbf{z}_t\mathbf{z}_t^T\right)^{-1}\left( \sum_{t=1}^n  \mathbf{z}_t x_t\right)
$$ given that matrix invertibility of the square matrix
$\sum_{t=1}^n \mathbf{z}_t\mathbf{z}_t^T$ (often written as $X^T X$ with
design matrix $X=(\mathbf{z}_1,\ldots,\mathbf{z}_n)^T$) holds.

We write $$
SSE=Q(\hat{\boldsymbol{\beta}}),
$$ the sum of squared errors after regression.

One then estimates the variance $\sigma_w^2$ of $\{w_t\}$ by $$
\hat{\sigma}_w^2=\frac{SSE}{n-(q+1)}=MSE.
$$

The variance-covariance matrix of $\hat{\boldsymbol{\beta}}$ is given by
$$
\sigma_w^2 C,
$$ where $C=\left(\sum_{t=1}^n \mathbf{z}_t\mathbf{z}_t^T\right)^{-1}$.

Under the assumption that the linear regression model is correct and
$\{w_t\}$ is Gaussian white noise, one can derive that\
$$
\frac{\hat{\beta}_i-\beta_i}{\hat{\sigma}_w \sqrt{C_{ii}}}\sim   \text{t-distribution with d.f. }n-q-1.
$$ This can be used to carry out t-test of $H_0: \beta_i=0$ vs
$H_a: \beta_i\neq 0$.

**Note.** However as mentioned above, Gaussian white noise assumption on
$\{w_t\}$ often fails for time series. So one should not read too much
into the result of the t-test in the time series context after
performing OLS.

**Example.**

In fact, often one can choose $z_{t,j}$ to be a function of time index
$t$ to model trends.

```{r}
#install.packages('astsa');
library('astsa')
```

The dataset *chicken* in *astsa*, which stores monthly price (cents per
pound) of a chicken in the US from mid-2001 to mid-2016 (180 months).

```{r}
chicken
length(chicken)
plot(chicken)
```

There is an obvious upward trend in the time series. Consider the
following model, suppose $x_t$ = chicken price.

$$
x_t=\beta_0+\beta_1 z_t + w_t,   \quad t=1,2,\ldots,180.
$$ where $$
z_1=2001+\frac{7}{12},\ z_2=2001+\frac{8}{12}, \ \ldots,\ z_{180}=2016+\frac{8}{12}.
$$

A convenient way to generate $\{z_t\}$ above is to directly extract it
from the *ts* object using the function *time*.

```         
time(x, ...)

time creates the vector of times at which a time series was sampled.
```

```{r}
z <- time(chicken)
attributes(z)
z
```

```{r}
summary(fit <- lm(chicken ~ z, na.action = NULL)) # na.action = NULL to retain the time series attributes for the residuals and fitted values.
```

```{r}
plot(chicken, ylab="cents per pound")
abline(fit) #plot fitted line.
```

```{r}
#another way to plot fitted value
ts.plot(chicken,fitted.values(fit),lty=1:2)
```

The estimated slope coefficient $\hat{\beta}_1 = 3.59$ (with a standard
error of 0.08) yielding a estimated increase of about 3.6 cents per
year. The t-test with extremely small p-value indicates strong evidence
of nonzero slope. However, the p-values should be treated with caution
since the residual exhibits strong temporal correlation. See below.

```{r}
plot(res<-residuals(fit)) # residual plot as time series
acf(res)
```

**Observation**: significant ACF values present. The model does not fit
well.

Exercise 21.

## 2.1.2 Variable Selection and Model Assessment

Recall the model

$$
x_t=\beta_0+\beta_1 z_{t,1}+\ldots+ \beta_q z_{t,q}+w_t. \tag{2}
$$

Often one may want to check if a simpler model involving fewer variables
is sufficient: $$
x_t=\beta_0+\beta_1 z_{t,1}+\ldots+ \beta_r z_{t,r}+w_t,  \quad r<q. \tag{3}
$$ Model (3) is obtained from (2) by setting
$\beta_{r+1}=\ldots=\beta_q=0$. We can then formulate the model
selection problem as testing the null hypothesis $$
H_0: \beta_{r+1}=\ldots=\beta_q=0.
$$ Let $SSE_r$ denote the sum of squared errors (after fitting OLS) of
the reduced model (3), and let $SSE$ still denote the sum of squared
errors of the full model (2). This is achieved through an F-test
involving the statistic $$
F=\frac{(SSE_r-SSE)/(q-r)}{SSE/(n-q-1)}=\frac{MSR}{MSE}
$$ which follows F distribution with d.f. $q-r$ and $n-q-1$ when the
null is true. We reject $H_0$ when the observed $F$ is large.

![](https://drive.google.com/uc?export=view&id=1HOcieJNkyshGOX34CvopBhmlb8Z2gMu7)

In particular, when $r=0$, we are testing
$H_0: \beta_1=\beta_2=\ldots=\beta_q=0$, under which the model becomes
$$
x_t=\beta_0+w_t,      
 $$ namely, no external explanatory variables are included. In this
case, the OLS estimate of $\beta_0$ is the sample mean
$\hat{\beta}_0 =\bar{x}$ and the sum of squared errors after fitting OLS
becomes $$
SSE_0=\sum_{t=1}^n (x_t-\bar{x})^2.
 $$ The following quantity $$
R^2=\frac{SSE_0-SSE}{SSE_0}
 $$ ($SSE$ is still regarding the full model) is called the *coefficient
of determination* which is an increasing function of $F$ statistic (see
[here](https://stats.stackexchange.com/questions/56881/whats-the-relationship-between-r2-and-f-test#:~:text=The%20practical%20interpretation%20is%20that,the%20coefficients%20are%20non%2Dzero.)).
So the larger $R^2$ is, the stronger evidence against
$H_0: \beta_1=\beta_2=\ldots=\beta_q=0$.

Below we use ANOVA (F-test) to compare the full model $$
x_t=\beta_0+\beta_1 z_t + w_t,   \quad t=1,2,\ldots,180.
$$ v.s. the reduced model $$
x_t=\beta_0+ w_t,   \quad t=1,2,\ldots,180.
$$

```{r}
summary(aov(fit))
```

ANOVA (F-test) can be used to add or delete variables in a stepwise
manner (one model is a submodel of the other).

For example, ANOVA can be used to compare model involving
$(1,z_{1,t},z_{2,t},z_{3,t})$ v.s. model involving $(1,z_{1,t})$, but
cannot be used to directly compare $(1,z_{1,t},z_{2,t})$ vs
$(1,z_{3,t})$.

Common measurements of goodness of fit/information criteria (smaller the
criterion better the fit):

-   Akaike’s Information Criterion (AIC)
-   AIC, Bias Corrected (AICc)
-   Bayesian Information Criterion (BIC)

They are all of the form $$
Criterion= \log(SSE(r)/n)+p_n(r)
$$ where $SSE(r)$ stands for $SSE$ resulting from regression on the
selected $r$ variables, and $p_n(r)$ is a penalty function which is an
increasing function of $r$. (Se Definitions 2.1-2.3 in the textbook for
more details.)

-   AICc $\approx$ AIC when sample size is large. AICc tends to be
    superior in smaller samples.

-   BIC tends to choose smaller models, does well at getting the correct
    order in large samples;

\*See Problems 2.4 and 2.5 in the textbook for more about these
information criteria (NOT homework).

**Example.** Monthly Pollution, Temperature and Mortality Data

```{r}
#install.packages('astsa');
library('astsa')
plot(cmort, main="Cardiovascular Mortality", xlab="", ylab="")
plot(tempr, main="Temperature", xlab="", ylab="")
plot(part, main="Particulates", xlab="", ylab="")

ts.plot(cmort,tempr,part, col=1:3)
```

```{r}
pairs(z<-cbind(Mortality=cmort, Temperature=tempr, Particulates=part)) # scatter plot matrix (notice renaming in cbind)
```

**Question**: does the plot matrix above contain temporal correlation
information?

```{r}
acf(z, lag.max = floor(length(cmort)*0.2))
```

Suppose

-   $M_t$: mortality
-   $T_t$: temperature
-   $P_t$: particulate level

Consider the following potential models $$
\begin{align}
M_t&=\beta_0+\beta_1t + w_t\\
M_t&=\beta_0 +\beta_1t +\beta_2 (T_t-T_{\bullet})+w_t\\
M_t&=\beta_0 +\beta_1t +\beta_2 (T_t-T_{\bullet})+\beta_3(T_t-T_{\bullet})^2 + w_t\\
M_t&=\beta_0 +\beta_1t +\beta_2 (T_t-T_{\bullet})+\beta_3(T_t-T_{\bullet})^2 + \beta_4 P_t+ w_t,
\end{align} 
$$

where $T_{\bullet}$ stands for mean of $\{T_t\}$ (adjusting for mean to
[avoid collinearity problem from polynomial
terms](https://stats.stackexchange.com/questions/183232/centering-variables-in-regression-leads-to-the-same-model-of-original-variables)).

```{r}
mean(tempr)
```

```{r}
temp = tempr-mean(tempr) # center temperature
temp2 = temp^2
trend = time(cmort) # time


fit = lm(cmort~ trend + temp + temp2 + part, na.action = NULL)

summary(fit) # regression results
```

Fitted value:

$$
\hat{M}_t=2832-1.396\times t -0.4725(T_t-74.26)+0.0259(T_t-74.26)^2 + 0.2554 P_t
$$

```{r}
# Change plot size
#options(repr.plot.width=20, repr.plot.height=7)

ts.plot(cmort, fitted.values(fit),lty=1:2,col=1:2)
legend('topleft', 'Fitted Value', col=2, lty = 2)

plot(res<-residuals(fit))
acf(res)
```

```{r}
summary(aov(fit)) # ANOVA table (foward step-wise F test)
```

Observation: each model does substantially better than the one before it

```{r}
summary(aov(lm(cmort~cbind(trend, temp, temp2, part)))) # ANOVA for overall regression H_0: beta_1=...beta_q=0
```

```{r}
num = length(cmort) # sample size
AIC(fit)/num - log(2*pi) # AIC
BIC(fit)/num - log(2*pi) # BIC
(AICc = log(sum(resid(fit)^2)/num) + (num+5)/(num-5-2)) # AICc
```

```{r}

fit_part = lm(cmort~ trend  + part, na.action = NULL)
fit_temp = lm(cmort~ trend + temp + temp2, na.action = NULL)


print('Information Criteria for model with part')
AIC(fit_part)/num - log(2*pi) # AIC
BIC(fit_part)/num - log(2*pi) # BIC
(AICc = log(sum(resid(fit_part)^2)/num) + (num+5)/(num-5-2)) # AICc


print('Information Criteria for model with temp and temp2')
AIC(fit_temp)/num - log(2*pi) # AIC (textbook version differs from R version by terms that do not change from model to model)
BIC(fit_temp)/num - log(2*pi) # BIC
(AICc = log(sum(resid(fit_temp)^2)/num) + (num+5)/(num-5-2)) # AICc
```

Exercise 22.

**Example.** Regression with lagged value.

$R_t$: Recruitment (fish population)

$S_t$: SOI

Model:

$R_t=\beta_0+\beta_1S_{t-6}+w_t$

Note:

for $S_{t-6}$, $t=7,\ldots,n+6$;\
for $R_t$, $t=1,\ldots,n$.

So only within the intersection $t=7,\ldots,n$ both are well-defined.

```         
ts.intersect(…, dframe = FALSE)
ts.union(…, dframe = FALSE)

Bind time series which have a common frequency. ts.union pads with NAs to the total time coverage, ts.intersect restricts to the time covered by all the series.
```

```{r}
soiL6=lag(soi,-6)
ts.plot(soi,soiL6,lty=1:2)


fish <- ts.intersect(rec, soiL6, dframe = T)
head(fish , 10)


summary(fit <- lm(rec~soiL6, data=fish, na.action=NULL))
```

```{r}
#options(repr.plot.width=16, repr.plot.height=6)
ts.plot(rec,fitted.values(fit),lty=1:2,col=1:2)
legend('topleft', 'Fitted Value', col=2, lty = 2)
plot(res<-residuals(fit))
acf(res,lag.max=60)
```

Exercise 23.
