---
title: An R Markdown document converted from "Section 1_5.ipynb"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# 1.5 Estimation of Mean, Covariance and Correlation for Stationary Time Series

## 1.5.1 Stationarity and Estimation

The mean and covariance/correlation introduced before are useful for
describing the properties of time series models.

However in practice, we typically only have sampled data which is
assumed to be generated by an unknown model. In such a setup, we do NOT
have direct access to these quantities. We can only **estimate** them
using data.

Suppose we have a time series $\{x_t\}$ with mean function $\mu(t)$ and
autocovariance function $\gamma(s,t)$.

Question: can we use a series of observed data, say
$\{x_1,\ldots, x_n\}$, to estimate $\mu(t)$ and $\gamma(s,t)$?

**Challenge**: $\mu(t)$ and $\gamma(s,t)$ may change as $s$ and $t$
changes. However, typically one has only a *single* observation $x_t$
corresponding to $\mu(t)$, and a *single* pair of observation
$(x_s,x_t)$ corresponding to $\gamma(s,t)$.

So in this case, unless some additional assumption is made, $\mu(t)$ and
$\gamma(s,t)$ cannot be estimated accurately.

This is where the *stationarity* assumption shines!

Under weak stationarity assumption, $\mu(t)=\mu$ is a constant, and
$\gamma(s,t)=\gamma(s-t)=\gamma(h)$ depends only on the lag $h=s-t$.

Suppose we have observed $\{x_1,\ldots, x_n\}$ from a weakly stationary
time series.

Then

-   $\mu$ is shared by every $\{x_1,\ldots, x_n\}$.
-   $\gamma(h)$ is shared by every pair
    $(x_{1+h},x_1), (x_{2+h},x_2),\ldots, (x_{n},x_{n-h})$.

Hence we have multiple data entries to estimate these quantities.

## 1.5.2 Sample Mean

**Definition.** Given the time series observations $\{x_1,\ldots,x_n\}$,
the *sample mean* is defined as $$
\bar{x}=\frac{1}{n}\sum_{t=1}^n x_t.
$$

Under the weak stationarity assumption
$$E(\bar{x})=\frac{1}{n}(\mu+\ldots +\mu)=\mu.$$

Sample mean is typically a good estimator of the constant mean $\mu$.
Under the weak stationarity assumption, one can derive an expression for
its variance\*:

$$
\begin{align*}
Var(\bar{x})&=Var\left(\frac{1}{n}\sum_{t=1}^n x_t\right)= Cov\left(\frac{1}{n}\sum_{s=1}^n x_s,\frac{1}{n}\sum_{t=1}^n x_t\right)\\
&=\frac{1}{n^2}\left( \sum_{s=1}^n \sum_{t=1}^n\gamma(s-t)\right)\\
&= \frac{1}{n^2}\left[  n \gamma(0) + (n-1)(\gamma(-1)+\gamma(1))+ (n-2)(\gamma(-2)+\gamma(2))+\ldots + (\gamma(1-n)+\gamma(n-1)) \right]\\
&= \frac{1}{n} \sum_{h=-n}^{n} \left( 1 - \frac{|h|}{n} \right)\gamma(h). \tag{1}
\end{align*}
$$

**Facts**

-   If $\{x_t\}$ is uncorrelated, then the formula (1) simplifies to $$
    \frac{\gamma(0)}{n}=\frac{\sigma_x^2}{n}, \tag{2}
    $$ a well-known expression for variance of sample mean of i.i.d.
    data.
-   Depending on the autocovariance function $\gamma(h)$, the variance
    of $\bar{x}$ may be smaller or greater than $\frac{\gamma(0)}{n}$:
-   if negative $\gamma(h)$ dominates, then $Var(\bar{x})$ is smaller
    than $\frac{\gamma(0)}{n}$ (rare situation);
-   if positive $\gamma(h)$ dominates, then $Var(\bar{x})$ is greater
    than $\frac{\gamma(0)}{n}$ (much more common situation).So often,
    accuracy of sample mean for time series is worse than that for
    i.i.d. data.
-   \*Under the assumption $\sum_h |\gamma(h)|<\infty$, one can show
    that as sample size $n\rightarrow\infty$, we have $$
    \sum_{h=-n}^{n} \left( 1 - \frac{|h|}{n} \right)\gamma(h) \rightarrow   \sum_{h=-\infty}^{\infty}  \gamma(h).
    $$ The quantity $\sigma^2= \sum_{h=-\infty}^{\infty} \gamma(h)$ is
    often called the long-run variance. So when $n$ is large (typically
    the case in time series analysis), $$
    Var(\bar{x})\approx \frac{\sigma^2}{n}. \tag{3}
    $$ Compare (2) and (3). Hence the long-run variance $\sigma^2$ is
    key for quantifying the accuracy of $\bar{x}$ (e.g., constructing
    confidence interval) when the sample size is large. Estimation of
    the long-run variance $\sigma^2$ is an important topic (explore in
    project).

**Example** Recall time series models: $x_t= w_{t}+w_{t-1}$ and
$y_t= w_t-w_{t-1}$, where $\{w_t\}$ is white noise with variance $1$.
Both are stationary linear processes with mean zero. We have derived
previously: $$
\gamma_x(h)=
\begin{cases}
2 &  h=0, \\
1 & |h|=1, \\
0 & |h|\ge 2,
\end{cases},
\qquad
\gamma_y(h)=
\begin{cases}
2 &  h=0, \\
-1 & |h|=1, \\
0 & |h|\ge 2,
\end{cases}
$$

One can use formula (1) to calculate $Var(\bar{x})$ and $Var(\bar{y})$.
Obviously $Var(\bar{y})< Var(\bar{z}) < Var(\bar{x})$, where $\{z_t\}$
is white noise with marginal variance $2$.

```{r}
xbar<-ybar<-zbar<-c()

for (i in 1:1000)
{

  x = na.omit(filter(rnorm(1001,0,1),c(1, 1)))
  y = na.omit(filter(rnorm(1001,0,1),c(1, -1)))
  z = as.ts(rnorm(1000,0,sqrt(2)))# note rnorm(n, mean = 0, sd = 1)

  xbar[i]=mean(x)
  ybar[i]=mean(y)
  zbar[i]=mean(z)
  # xbar ybar and zbar collect sample means realized from repetitive simulations of the three different models respectively
}
```

```{r}
ts.plot(window(x,1,50),window(y,1,50),window(z,1,50),col=c("blue","black","red"), lty=1:3) #plot simulated first 50 observations of the three models in the same plot
```

```{r}
var(xbar);var(zbar);var(ybar) # compute (sample) variance of simulated sample means for the 3 models
```

```{r}
#histograms of simulated sample means for the 3 models
par(mfrow=c(3,1))
hist(xbar,20)
hist(zbar,20)
hist(ybar,20)
```

Exercise 16. Explore more about mean estimation.

\
**Important Clarification**

When $\mu(t)$ is not a constant, (whole) sample mean typically cannot be
viewed as an estimate of certain population quantity. Still, sample mean
provides a useful summary of the location of the data points.

Sometimes even though $\mu(t)$ is a constant $\mu$ (but the time series
is NOT weakly stationary), the sample mean may NOT be a good estimate of
$\mu$.

**Example.** We know the mean function $\mu(t)=\mu=0$ for random walk
$x_t=w_1+\ldots+w_t$, where $\{w_t\}$ is standard Gaussian white noise.
However, $\bar{x}$ is NOT a good estimate of $\mu=0$.

```{r}
plot(x<-filter(rnorm(5000),filter=1,method = 'recursive')) #simulate random walk starting at zero
abline(h=mean(x)) # draw sample mean as a horizontal level
```

## 1.5.3 Sample Autocovariance and Autocorrelation

Recall if one has $m$ pairs of data points
$\{(u_1,v_1),(u_2,v_2),\ldots, (u_m,v_m)\}$, one can compute the sample
covariance as $$
\frac{1}{m}\sum_{i=1}^m (u_i -\bar{u}) (v_i-\bar{v})   \tag{4}
$$ where $\bar{u}=\frac{1}{m}\sum_{i=1}^m u_i$,
$\bar{v}=\frac{1}{m}\sum_{i=1}^m v_i$.

If $\{(u_i,v_i)\}$ are randomly drawn from the bivariate distribution of
the random vector $(U,V)$, then the sample covariance is an estimator of
$Cov(U,V)=E[(U-E(U))(V-E(V))]$.

Setting $v_i=u_i$ in the formula (4) recovers the sample variance for
$\{u_1,\ldots,u_m\}$: $$
\frac{1}{m}\sum_{i=1}^m (u_i -\bar{u})^2. \tag{5}
$$ which is an estimator of the variance of $U$.

**Remark.** Often people replace the factor $\frac{1}{m}$ in (4) and (5)
by some other factor such as $\frac{1}{m-1}$. It makes little difference
when $m$ is large. We will also use a factor sometimes different from
$\frac{1}{m}$ later.

We also have the sample correlation $$
\frac{\sum_{i=1}^m (u_i -\bar{u}) (v_i-\bar{v})}{\sqrt{\sum_{i=1}^m (u_i -\bar{u})^2} \sqrt{\sum_{i=1}^m (v_i-\bar{v})^2}}.
$$ which is an estimator of $Corr(U,V)$.

```{r}
#install.packages('MASS');
library('MASS')
```

```{r}
v=2;rho=0.9 # specify marginal variance and correlation
z = mvrnorm(500, mu=c(0,0), Sigma=matrix(c(v,rho*v,rho*v,v), nrow = 2, ncol = 2)) #generate sample from bivariate normal distribution with marginal variance v (same for both components)  and correlation rho.
head(z,10) # show first 10 sample pairs
plot(z)
u=z[,1];v=z[,2]
print('sample variance')
var(u) # by default R computes sample variace/covariance by dividing by (sample size-1)
print('sample covariance')
cov(u,v)
print('sample correlation')
cor(u,v) # compute sample correlation
```

Now given sample $\{x_1,\ldots,x_n\}$ from a weakly stationary time
series, we want to estimate the autocovariance/correlation functions
$\gamma(h)$ and $\rho(h)$.

Again recall by weak stationarity, all the pairs
$(x_{1+h},x_1), (x_{2+h},x_2),\ldots, (x_{n},x_{n-h})$ share the same
(theoretical/population) covariance value $\gamma(h)$.

So one can use the sample covariance formula (4) to estimate
$\gamma(h)$, i.e., simply replace $u_i$ by $x_{i+h}$, and $v_i$ by $x_i$
there.

**Definition.** Given time series observations $\{x_1,\ldots,x_n\}$, the
**sample autocovariance function** is defined as $$
\hat{\gamma}(h)=\frac{1}{n}\sum_{t=1}^{n-h}(x_{t+h}-\bar{x})(x_t-\bar{x})
$$ where $\bar{x}=(x_1+\ldots+x_n)/n$ is the sample mean.

***Remark: it is a convention to use*** $\frac{1}{n}$ instead of
$\frac{1}{n-h}$ in the definition.

Reasons:

-   (Most important one\*) The choice of $\frac{1}{n}$ makes
    $\hat{\gamma}(h)$ non-negative definite (consistent with the same
    property of population/theoretical autocovariance function);

-   When $|h|$ is small compared to $n$,
    $\frac{1}{n} \approx \frac{1}{n-h}$, so the difference is
    negligible;

-   When $|h|$ is close to $n$, the sum $\sum_{t=1}^{n-h}$ has few terms
    and the estimate is inaccurate. The choice $\frac{1}{n}$ (instead of
    $\frac{1}{n-h}$) tends to make $\hat{\gamma}(h)\approx 0$ for $|h|$
    close to $n$, which is often the case for the theoretical
    $\gamma(h)$.

**Definition** Given time series observations $\{x_1,\ldots,x_n\}$ and
sample covariance function $\hat{\gamma}(h)$ above, the **sample
autocorrelation function (ACF)** is defined as $$
\hat{\rho}(h)= \frac{\hat{\gamma}(h)}{\hat{\gamma}(0)}, \quad |h|=0,1,\ldots, n-1.
$$

***In practice, people usually only examine sample autocorrelation
function (sample autocovariance function takes the same shape).***

```{r}
#install.packages('astsa');
library('astsa')
```

```{r}
attributes(soi) # show attributes of the object soi.
plot(soi) # climate index
```

In R, the computation of sample autocovariance/correlation is
implemented through the function
'[acf](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/acf)'.

```         
The function acf computes (and by default plots) estimates of the autocovariance or autocorrelation function.

acf(x, lag.max = NULL,
    type = c("correlation", "covariance", "partial"),
    plot = TRUE, na.action = na.fail, demean = TRUE, …)


Value
An object of class "acf", which is a list with the following elements: ...
```

```{r}
attributes(soi_acf<- acf(soi,plot = F))
```

```{r}
soi_acf$acf # ACF values at lag 0,1,2,..., lag.max  (if left unspecified, is given by a formula of sample size).
soi_acf$acf[2]
```

```{r}
(r <- soi_acf$acf[-1]) # store ACF values (delete the first element which corresponds to lag 0).
```

```{r}
soi_acf$type # which type, e.g., covariance or correlation
```

```{r}
soi_acf$lag # the lags at which (can be non-integers) the acf is estimated
```

```{r}
soi_acf$series # name of original time series
```

```{r}
soi_acf$n.used # The number of observations in the time series
```

The **lag.plot** plots a scatter plot of data points
$\{(x_{t+h},x_t),t=1,\ldots n-h \}$ which helps visualizing
auto-dependence.

```{r}
lag.plot(soi,set.lags=1) #set.lags = h
legend('topleft', legend=c('correlation',r[1]))
lag.plot(soi,set.lags=5)
legend('topleft', legend=c('correlation',r[5]))
```

Another important plot is the plot of ACF values vs lag. An 'acf' object
admits a special method of plot.

```{r}
acf(soi,lag.max = 50)
```

**Question**: *What do the dashed lines in the acf plot standard for?*

```{r}
acf(rnorm(1000),lag.max = 100) #sample ACF for Gaussian white noise
```

Exercise 17. Explore this in depth. (Knowledge of the concept of
confidence interval is helpful)

**Note**: For a weakly stationary time series, sample ACF can be viewed
as an estimate of the theoretical ACF.

Exercise 18. Sample ACF vs Theoretical ACF for moving average.

\
**Important Clarification**:

-   Theoretical **univariate** autocovariance/correlation $\gamma(h)$
    and $\rho(h)$ makes sense only when the the time series model is
    weakly **stationary**.

-   the sample autocovariance/correlation $\hat{\gamma}(h)$ and
    $\hat\rho(h)$ can always be computed given an observed time series
    $\{x_1,\ldots,x_n\}$, regardless of whether it satisfies weak
    stationarity assumption or not.

-   However when weak stationarity does not hold, $\hat{\gamma}(h)$ and
    $\hat\rho(h)$ often CANNOT be viewed as estimates of certain
    theoretical values.

-   Still for any observed time series (stationary or not), sample ACF
    provides a useful summary of the temporal correlation.

$\bullet$ Random walk

We know that the random walk model $x_t=w_1+\ldots+w_t$, where $\{w_t\}$
is standard Gaussian white noise, is NOT weakly stationary.

Recall $x_t=x_{t-1}+w_t$

```{r}
plot.ts(x<-filter(rnorm(5000) , filter=1 , method="recursive") , main="random walk")
```

```{r}
acf(x)
```

\*One can theoretically verify that as sample size increases, the sample
correlation function $\hat{\rho}(h)$ of a random walk converges to $1$
(in probability) at any lag $h$ (see Exercise 1.8 in the textbook, NOT
part of homework).

$\bullet$ Speech time series

```{r}
plot(speech)

acf(speech, 250)
```

The highest peak (except at lag 0) in sample ACF above reflects
periodicity (about 106-109 time points) of the original speech signal.

Although the speech siginal is far from stationary (e.g., may not have a
constant mean function), ACF still provides useful information.

Exercise 19 Sample ACF for signal-plus-noise model.

## 1.5.4 Sample Cross-Covariance and Cross-Correlation

The concepts of sample mean and covariance/correlation for univariate
time series can be naturally extended to multivariate cases.

**Definition.** Suppose we have $r$-dimensional time series observations
$\{\mathbf{x}_1,\ldots, \mathbf{x}_n \}=\{(x_{1,1},\ldots, x_{1,r})^T,\ldots, (x_{n,1},\ldots, x_{n,r})^T\}$
of sample size $n$.

-   The **sample mean vector** is given by\
    $$\bar{\mathbf{x}}=(\bar{x}_1,\ldots,\bar{x}_r),
    $$ where $\bar{x}_j=(x_{1,j}+\ldots+x_{n,j})/n$ stands for the
    sample mean of jth component.
-   The **sample cross-covariance function** is defined as $$
    \hat{\gamma}_{j,k}(h)=\frac{1}{n}\sum_{t=1}^{n-h}(x_{t+h,j}-\bar{x}_j)(x_{t,k}-\bar{x}_k).
    $$
-   The **sample cross-correlation function** is defined as $$
    \hat{\rho}_{j,k}(h)=\frac{\hat{\gamma}_{jk}(h)}{\sqrt{ \hat{\gamma}_{jj}(0)\hat{\gamma}_{kk}(0) }}.
    $$

```         
ccf(x, y, lag.max = NULL, type = c("correlation", "covariance"),
    plot = TRUE, na.action = na.fail, …)


The lag k value returned by ccf(x, y) estimates the correlation between x[t+k] and y[t].
```

```{r}
#install.packages('astsa');
#library('astsa')
plot(soi,main="Southern Oscillation Index")
plot(rec,main="Recruitment")
```

```{r}
acf(soi, 48, main="Southern Oscillation Index")
acf(rec, 48, main="Recruitment")
ccf_sr<-ccf(soi, rec, 48, main="SOI vs Recruitment", ylab="CCF")
ccf(rec, soi, 48, main="SOI vs Recruitment", ylab="CCF")
```

**Warning**: unlike in ACF plot, the dashed lines in CCF are NOT readily
applicable unless one of the time series is approximately white noise.

(To make the dashed line work, it involves a procedure called
"pre-whitening". See Property 1.3 and Example 1.29 in the textbook\*).

```{r}
attributes(ccf_sr)
```

```{r}
length(ccf_sr$acf)

ccf_sr$acf[(length(ccf_sr$acf)+1)/2] # the cross correlation corresponding to lag h = 0
```

```{r}
plot(lag(soi,1),rec) #lag h = 1
legend('topleft',legend=round(ccf_sr$acf[(length(ccf_sr$acf)+1)/2 +1],4)) # round up to 4 digits when showing ccf in legend

plot(lag(soi,-6),rec) #lag h = -6
legend('topleft',legend=round(ccf_sr$acf[(length(ccf_sr$acf)+1)/2 -6],4))
```

**Observation**: SOI measured at time (around) $t-6$ months is
negatively associated with the Recruitment series at time $t$.

It is often useful to use matrix notation to summarize (sample)
cross-covariance/correlation.

**Definition.** For a $r$-dimensional weakly stationary time series
$\{\mathbf{x}_t=(x_{t,1},\ldots,x_{t,r})^T\}$ with mean vector
$\boldsymbol{\mu}=(\mu_1,\ldots,\mu_r)^T$ and cross-covariance functions
$\gamma_{i,j}(h)$, we define the **autocovariance matrix** as $$
\Gamma(h)=[\gamma_{i,j}(h)]_{i,j=1,\ldots,r}=
\begin{bmatrix}
\gamma_{1,1}(h), \gamma_{1,2}(h),\ldots, \gamma_{1,r}(h)\\
\gamma_{2,1}(h), \gamma_{2,2}(h),\ldots, \gamma_{2,r}(h)\\
\ldots \\
\gamma_{r,1}(h), \gamma_{r,2}(h),\ldots, \gamma_{r,r}(h)
\end{bmatrix}
  =E [(\mathbf{x}_{t+h}-\boldsymbol{\mu}) (\mathbf{x}_t-\boldsymbol{\mu})^T]
$$ where $E$ above is taken element-wise with respect to the matrix $$
(\mathbf{x}_{t+h}-\boldsymbol{\mu}) (\mathbf{x}_t-\boldsymbol{\mu})^T=[(x_{t+h,i}-\mu_i)(x_{t,j}-\mu_j)]_{i,j=1,\ldots,r}
$$ **Autocorrelation matrix** is defined similarly with
cross-correlations replacing the entries of the matrix.

**Definition.** Suppose we have $r$-dimensional time series observations
$\{\mathbf{x}_1,\ldots, \mathbf{x}_n \}=\{(x_{1,1},\ldots, x_{1,r})^T,\ldots, (x_{n,1},\ldots, x_{n,r})^T\}$
of sample size $n$. The **sample autocovariance matrix** is defined as
$$
\hat{\Gamma}(h)=[\hat{\gamma}_{i,j}(h)]_{i,j=1,\ldots,r}=\frac{1}{n}\sum_{t=1}^{n-h} (\mathbf{x}_{t+h}-\bar{\mathbf{x}}) (\mathbf{x}_t-\bar{\mathbf{x}})^T
$$ where $\hat{\gamma}_{i,j}(h)$ are sample cross-covariance functions
between $i$th and $j$th component with lag $h$ (time of component $i$
minus time of component $j$) as before.

**Sample autocorrelation matrix** is defined similarly with sample
cross-correlations replacing the entries of the matrix.

```{r}
attributes(x<-cbind(soi,rec)) #combining soi and rec into a bivariate time series x
x[1:10,1]==soi[1:10]#check whether first colum of x is soi (first 10 observations)
```

```         
ts(data = NA, start = 1, end = numeric(), frequency = 1,
   deltat = 1, ts.eps = getOption("ts.eps"), class = , names = )
   
data
a vector or matrix of the observed time-series values.

class
class to be given to the result, or none if NULL or "none". The default is "ts" for a single series, c("mts", "ts", "matrix") for multiple series.
```

```{r}
attributes(x_acf<-acf(x,lag.max = 48)) # the function 'acf' now essentially computes sample autocorrelation matrix of x
```

Recall the relation $\gamma_{ij}(h)=\gamma_{ji}(-h)$. Similar relation
also holds for (sample) cross-correlations.

```{r}
 dim(x_acf$acf)

print('lag 0')
 x_acf$acf[0+1,,] #sample autocorrelation matrix at lag h=0

print('lag 6')
x_acf$acf[6+1,,] # at lag h=6
```

So in a sense the functionalities of *acf* covers those of *ccf* since
the former can also compute cross-correlations, while the latter may be
convenient at times.

Exercise 20

\*Interested students are encouraged to read Section 1.6 of the
textbook, which addresses the case where the "time index" is
multi-dimensional (in contrast to the value of each observation being a
vector for multivariate time series).
