---
title: "Section 1_1"
output:
  html_document:
    df_print: paged
---

**Definition** Time series: a series of data points indexed in time order.

Classically, there are two main approaches (not mutually exclusive) in time series analysis:

-   **Time domain approach**: focus on investigation of lagged relationships (e.g., how does what happened today affect what will happen tomorrow?).

-   **Frequency domain approach**: focus on investigation of cycles (e.g., what is the economic cycle through periods of expansion and recession?).

We will mainly address the time domain approach in the course.

# 1.1 The Nature of Time Series Data

```{r}
#install.packages("astsa");
library(astsa) 
```

**Example** Johnson & Johnson Quarterly Earnings

```{r}
jj
```

```{r}
class(jj) 
```

```{r}
plot(jj, type="o", ylab="Quarterly Earnings per Share")
```

**Observations**: exponential trend, cycle, random fluctuations.

Questions we can ask:

-   Predict next quarter earning.
-   Find external factors that explain earning fluctuation

...

**Example** Temperature anomalies

Annual temperature anomalies (in degress centigrade) averaged over the Earth's land and ocean area from 1850 to 2023. Anomalies are with respect to the 1991-2020 average.

```{r}
gtemp_both                 
```

```{r}
plot(gtemp_both, type="o", ylab="Global Temperature Deviations")
```

Observations

-   Irregular (random) fluctuation
-   Potential trend?

Questions we can ask:

-   Is there a breakpoint (or multiple ones) in the trend? $\Rightarrow$ Global warming?

(Project topic: [Change Point Detection](https://en.wikipedia.org/wiki/Change_detection)).

**Example**: Speech Signal

100ms-frequency sample of recorded speech for the phrase aaa $\cdots$ hhh.

```{r}
speech
```

```{r}
plot(speech)
```

Observations

-   Regular repetition of cycles (*pitch periods*).
-   Amplitude changes over time while cyclic pattern is relatively stable.

Questions we can ask:

-   How can one recognize pattern in frequency domain?

(Topic: [Spectral Analysis](https://en.wikipedia.org/wiki/Spectral_analysis))

Our course will NOT go deep into analysis of regular cyclic time series (specially the ones with large-sample and high-frequency). E.g., electric/sound/light signals.

Refer to the subject [Signal Processesing](https://en.wikipedia.org/wiki/Signal_processing)

Statistical time series analysis emphasizes on analysis of time series *exhibiting irregular/random fluctuations*

```{r}
spec.pgram(speech) # wait until Chapter 4 to see
```

**Example**: Dow Jones Industrial Average

```{r}
#install.packages("xts");
library(xts)
```

```{r}
class(djia)
```

```{r}
djia[,4]
```

```{r}
head(djia,10) # first 10 rows of data
plot(djia$Close)
```

Suppose $x_t$ is the closing price at $t$th day.

However, people are usually more interested in the relative price change:

$\tilde{r}_t= \frac{x_t-x_{t-1}}{x_{t-1}}$.

In finance, people also look at a quantity called *log return*:

$r_t=\log(x_t/x_{t-1})=\log(x_t)-\log(x_{t-1})=\log\left(1+ \frac{x_t-x_{t-1}}{x_{t-1}} \right)\approx \tilde{r}_t$.

Fact: $\log(1+y)\approx y$ when $|y|$ is small.

```{r}
curve(log(1+x),-0.2,0.2) #plot log(1+x) function
abline(h=0, v=0)#plot x-y axes
```

**Definition** The operation $(x_1,x_2,\ldots,x_n)\rightarrow (x_{\ell+1}-x_{1}, x_{\ell+2}-x_2,\ldots, x_{n}-x_{n-\ell})$ is called differencing, where $\ell=1,2,3,\ldots$ is called the *lag*. We take the lag $\ell=1$ by default if not specified otherwise.

In R it can be achieved with [diff](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/ts-methods) function.

```{r}
x<-c(1,3,5,7,9)
(diff(x))
(y<-diff(x,lag=2))
```

We first plot the log-transformed closing price.

```{r}
plot(djial<-log(as.ts(djia[,4])))
```

Log return can be obtained by first take log elementwise and then perform differencing (lag=1).

```{r}
djiar = diff(as.ts(djial))  #log return
plot(djiar, main="DJIA Returns")
```

Observations

-   Clustering of fluctuation magnitude (volatility clustering)

```{r}
acf(djiar) # compute & plot sample auto-correlation function (will elaborate later in the course) for log return
acf(abs(djiar)) # compute & plot sample auto-correlation function for  absolute value of log return
```

Project topic: GARCH models.

**Example** El Niño and Fish Population

Monthly values of an environmental series called the Southern Oscillation Index (*SOI*, changes in air pressure) and associated *Recruitment* (population of new fish).

```{r}
plot(soi, ylab="", xlab="", main="Southern Oscillation Index")
plot(rec, ylab="", xlab="", main="Recruitment")
```

Observations

-   Both collected at the same time points.
-   Annual cycle and 4-year cycle.
-   Two series are related.

Questions we can ask:

-   Are the cycles changing over time?
-   Joint analysis of two time series. E.g., do SOI and Recruitment have a "causal" relationship?

Time-Frequency Analysis (e.g., [Wavelet](https://en.wikipedia.org/wiki/Wavelet))

Causal analysis (e.g., [Granger Causality](https://en.wikipedia.org/wiki/Granger_causality))

**Exercise 6.** Plot multiple time series in a single graph.
